Glyph-maps for visualizing climate data and models

### 2011-08-29 paper.tex
references
finished tweaking intro

In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.  For weather maps, icon plots go back as far as \citet{galton:weathermap}. Modern icon plots date back to the semi-graphic display of \citet{anderson:1960} and star glyphs of  \citep{kleiner:1981}, and were made infamous by \citep{chernoff:1973}. An icon is produced by mapping each variable to some feature of the icon. One icon is generated for every case in the data set. Figure \ref{fig:glyphs} illustrates the idea of icon plots, with emphasis on temporal data. The time measurements are treated as variables in a multivariate data set. Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables. Three different locations are selected, for demonstration purposes. The top plot shows the time series of these three locations. Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six peaks. Location ``2.1'' (green) has generally high temperature values, with weaker seasonality in most years and a lack of low temperatures in the middle of the time period when lows would be expected. Location ``3.1'' (blue) has strong seasonality but generally low temperatures. The bottom plot shows these same cases as star glyphs. A separate glyph is created for each location. The 72 time variables form axes organized radially. Each temperature value is plotted relative to the overall range of temperature values. The seasonal patterns are clearly visible in this type of display, forming six ``petals'' in the icon. Location ``2.1'' has a different pattern than the other two, with only four petals, and location ``3.1'' is a much smaller icon indicating overall lower values.

@@

Icon plots are used, infrequently, to examine multivariate data. For large data sets it is infeasible to digest the quantity of icons, and for multivariate data there is no natural order so it can be like having a mountain of little stamps to rearrange and organize, in order to understand the structures present in the data. The initial paper on star glyphs \citep{kleiner:1981} and a recent paper \citep{hurley:2010} propose some approaches for organizing the icons. With spatiotemporal data ordering of icons is naturally done by spatial location. There is also usually some spatial trend, or continuity, which will make digesting patterns among the icons simpler.

@@

Glyph-maps have been used in

This paper describes how to make glyph-maps.

@@

Code and data is available in the online supplementary materials.

\section{Computation}

@@

\section{Reference lines}

@@

\section{Generalization: scatterplots}

\section{Partitioning patterns into shape and size}

@@

\section{Regular vs irregular locations}

>>>>

%% Reproduce all of the icon displays using the fixed glyph code

In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.  Icon plot associated with weather maps date back as far as \citet{galton:weathermap}, described in \citet{friendlydenis:2001}. Icon plots date back to the star plot introduced by \citet{mayr:1877}, and resurged in the modern era with the semi-graphic displays of \citet{anderson:1960}, and were made infamous by \citep{chernoff:1973}. An icon is produced by mapping each variable to some feature of the icon. One icon is generated for every case in the data set. Figure \ref{fig:glyphs} illustrates the construction of star plot, with emphasis on temporal data. The time measurements are treated as variables in a multivariate data set. Variables are Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables. The star plot is formed by considering the 72 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable. For the temporal data, all variables are considered to be on a common scale so the global minimum and maximum of temperature are used to scale each data value. Three different locations are selected, for demonstration purposes. The top plot shows the time series of these three locations. Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six peaks. Location ``2.1'' (green) has generally high temperature values, with weaker seasonality in most years and a lack of low temperatures in the middle of the time period when lows would be expected. Location ``3.1'' (blue) has strong seasonality but generally low temperatures. The bottom plot shows these same cases as star glyphs. A separate glyph is created for each location. The 72 time variables form axes organized radially. Each temperature value is plotted relative to the overall range of temperature values. The seasonal patterns are clearly visible in this type of display, forming six ``petals'' in the icon. Location ``2.1'' has a different pattern than the other two, with only four petals, and location ``3.1'' is a much smaller icon indicating overall lower values.

@@

Icon plots are used, infrequently, to examine multivariate data. For large data sets it is infeasible to digest the quantity of icons, and for multivariate data there is no natural order so it can be like having a mountain of little stamps to rearrange and organize, in order to understand the structures present in the data. An early paper on multivariate graphics \citep{kleiner:1981} and a recent paper \citep{hurley:2010} propose some approaches for organizing the icons. Ordering of icons is not an issue with spatiotemporal data, because it is naturally handled by spatial location. The spatial trend, or continuity, usually gices the plot an appearance of texturing on a landscape, make it easier to digest patterns among the icons.

@@

The next section (\ref{sec:construction}) describes how to construct glyph-maps. Section \ref{sec:reference} discusses the use of reference lines and boxes, and Section \ref{sec:scale} visits issues of scaling values for different information purposes. Generalizations to different types of icons are described in Section \ref{sec:scatter}. Glyph-maps work nicely with gridded data, but may be less effective with irregular spatial locations. Possible adjustments are discussed in Section \ref{sec:irregular}. Code and data is available in the online supplementary materials.

@@

\section{Computation}~\label{sec:construction}

@@

\section{Reference lines}~\label{sec:reference}

@@

\section{Partitioning patterns into shape and size}~\label{sec:scale}

@@

\section{Generalization: scatterplots}~\label{sec:scatter}

\section{Regular vs irregular locations}~\label{sec:irregular}
